### 2019-01-06 paper.Rmd
add paper outline

---
title: "R Markdown for Reproducible Research"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# ABSTRACT

R Markdown is one of many tools available to the R programmer when it comes to
building and sharing the modeling and data pipeline work that has been written
in R. The ability to reliably reproduce results is essential to the success and
impact of a researcher’s work. We will show that R Markdown is a valuable tool
for this task by introducing R Markdown, showing a handful of examples, and
discussing best practices. When all prose, reporting, and output is generated by
a single file based on data inputs, the researcher is freed from semantics and
can enjoy truly data-driven results.

# INTRODUCTION

# GETTING STARTED

## WHAT IS MARKDOWN?

## WHAT IS R MARKDOWN?

## WHAT IS AN R NOTEBOOK?

## WHAT ABOUT OTHER LANGUAGES?

## OUTPUT TYPES

## PARAMETERIZED R MARKDOWN

# A RESEARCHER'S WORKFLOW

# OTHER TOOLS

# CONCLUSION

### 2019-01-06 paper.Rmd
add intro and resources

@@
 
+The researcher's task is to uncover new things; challenge conventional wisdom;
+explore uncharted territory; and confirm or dismiss bias with facts. In order to
+accomplish such feats, it is important that a researcher's results be
+_reproducible,_ that is:
+
+> Results are reproducible if they can be reliably replicated or recomputed in
+another time or place.
+
+For a researcher that is trying to have an impact that extends beyond the "here and now,"
+reproducible results ensure that conclusions are valid in another time and place. Further,
+for industries like the Pharmaceutical Industry, reproducibility can be a mandate and 
+legal requirement for valid results. As a result, tools like R Markdown that help ensure
+reproducible results are valuable tools in a researcher's toolkit.
+
 # GETTING STARTED
@@
 
+The story of R Markdown begins with Markdown, a markup language and formatting
+syntax designed to be:
+
+ - human readable
+ - flexible in its output
+
+For example, where HTML would allow writing bold and italicized text like:
+
+```
+This is <b>bold</b> and <i>italicized</i> text
+```
+
+Markdown allows bold and italicized text with:
+
+```
+This is **bold** and _italicizied_ text
+```
+
+A subtle difference, but Markdown is more readable because there are not any
+letters injected, and it is _much faster_ to type. It also has more simplicity
+in its objectives and flexibility in its output, so where HTML is rendered
+almost exclusively in browsers, Markdown can be relevant for the web, print, and
+even computer terminals.
+
 ## WHAT IS R MARKDOWN?
+
+If Markdown is a lightweight markup language and text formatting syntax, R Markdown
+must be Markdown + R code, right? Yes, with a catch. R Markdown actually supports
+_many_ programming languages and output types. It got its start with R, and R is
+the primary language engine, but R is an interface language. This means that R
+is commonly used to "wrap" or
 
@@
 # CONCLUSION
+
+### Resources
+
+- https://rpubs.com/marschmi/105639
+- http://ropensci.github.io/reproducibility-guide/sections/tools/
+- http://shop.oreilly.com/product/0636920051435.do
+- https://rviews.rstudio.com/2017/03/15/why-i-love-r-notebooks/
+- https://rviews.rstudio.com/2018/11/01/r-markdown-a-better-approach/
+- https://rviews.rstudio.com/2017/01/25/r-markdown-for-the-enterprise/
+- https://rviews.rstudio.com/2018/01/18/package-management-for-reproducible-r-code/
+- https://en.wikipedia.org/wiki/Markdown
+- https://www.oreilly.com/library/view/the-art-of/9781593273842/ch15.html

### 2019-01-06 paper.Rmd
add explanatory r markdown section

@@
 
-If Markdown is a lightweight markup language and text formatting syntax, R Markdown
-must be Markdown + R code, right? Yes, with a catch. R Markdown actually supports
-_many_ programming languages and output types. It got its start with R, and R is
-the primary language engine, but R is an interface language. This means that R
-is commonly used to "wrap" or
+If Markdown is a lightweight markup language and text formatting syntax, R
+Markdown must be Markdown + R code, right? Yes, with a catch. R Markdown
+actually supports _many_ programming languages and output types. It got its
+start with R, and R is the primary language engine, but R is an interface
+language. This means that R has been (from its outset) commonly used to "wrap"
+or call other programming languages. This began as Fortran and C, but has
+expanded to include C++, Python, JavaScript, SQL, Stan, and more. As a result, R
+Markdown supports the inclusion of many of these programming languages.
+
+## THE YAML HEADER
+
+We will focus on R code for now, and come back to the other languages later. R
+Markdown has a handful of noteworthy differences from general Markdown. First,
+every R Markdown document begins with a YAML header:
+
+```
+---
+title: "R Markdown for Reproducible Research"
+output: pdf_document
+---
+```
+
+The YAML header uses YAML syntax to declare all sorts of information about the
+document. The full documentation on what options are available is [here](todo),
+but the primary elements are the document's Title and Output type. The output type
+has many options, and can actually support multiple options at the same time.
+
+Some choices for Output Type:
+
+- Static HTML page
+- HTML website
+- PDF document
+- Microsoft Word document
+- HTML Presentation
+- PDF Presentation
+- Microsoft PowerPoint Presentation
+- Shiny (interactive) web application
+- Blog, Book, etc.
+
+## THE REST OF THE DOCUMENT
+
+Once the YAML header is defined, the body of the document contains:
+
+- Markdown text
+- inline code
+- code chunks
+- embedded items (images, etc.)
+
+Inline code is written like ` r my_function ` and gets placed into the document
+like text. It can be very useful for displaying data-determined values
+programmatically so that they are dynamically changed when the document is
+regenerated (for instance, if new data is acquired).
+
+Code chunks are much like inline code, but they stand alone and are more useful for
+lengthier code blocks that do complex logic, setup, value preparation, plotting, or
+other output generation.
+
+Embedded images and other resources are included into R Markdown much like they
+would be in an HTML page. The syntax looks like `![Text alias](path/to/resource)`.
+
+Once the document is written, it is passed through the `knitr` package to "knit"
+the R Markdown document into its final output(s). This creates a new
+/ fresh R context to render the document and thereby helps ensure
+reproducibility.
 
@@
 
-## WHAT ABOUT OTHER LANGUAGES?
-
-## OUTPUT TYPES
+## WHAT ABOUT THE OTHER LANGUAGES?
 
@@
 - https://www.oreilly.com/library/view/the-art-of/9781593273842/ch15.html
+
+### TODO
+- Link for YAML syntax
+- Link for YAML header docs
+- Resource on R as an interface language

### 2019-01-06 paper.Rmd
start section on r notebooks

@@
 lengthier code blocks that do complex logic, setup, value preparation, plotting, or
-other output generation.
+other output generation. It is worth noting that code chunks are [very flexible](todo)
+in their options and defaults, and the number of options can sometimes be overwhelming.
 
@@
 
+When creating a new document in the RStudio IDE, you might notice an option
+to create a "R Notebook" or "R Markdown Notebook." This is offered in addition
+to the long-known "R Markdown," and the differences are subtle.
+
+![RStudio IDE New File Dropdown showing R Markdown and R Notebook
+options](./img/new-file-dropdown.png)
+
 ## WHAT ABOUT THE OTHER LANGUAGES?
@@
 # OTHER TOOLS
+
+# TIPS AND TRICKS
+
+- Code chunk options
+- Knit options
+- URLs and internal links
+- RStudio documentation, etc
 

### 2019-01-06 paper.Rmd
add a few todo items
add a few tips and tricks

@@
 - URLs and internal links
+- Table of Contents
+- Table Output
 - RStudio documentation, etc
@@
 - Resource on R as an interface language
+- link to figures / images
+- start managing references better...

### 2019-01-07 paper.Rmd
finish up section on r notebooks

@@
 
-```
+```yaml
 ---
@@
 
+The most noteworthy difference about R Notebooks versus plain R Markdown is that
+they are somewhat _stateful_. This means that they cache and display output as
+it is generated, making it easier to interactively and iteratively build a
+narrative without having to re-render the document each time you have made a
+change.
+
+It is worth noting that this output can be **problematic for reproducibility**,
+though, since you may have unknowingly persisted an output (state) that is no
+longer relevant or reproducible.
+
+As a result, R Notebooks can be thought about as good for interactive
+exploration or discussion among developers, but moving to an R Markdown format
+is generally preferred once the report has stabilized or is being shared with an
+end user.
+
+R Notebooks are identified by `output_type: html_notebook`. Remember that you can have
+multiple output types, so a YAML header like the following allows you to easily
+build multiple output types from the same document.
+
+```yaml
+---
+title: "R Notebook"
+output:
+  html_document:
+    df_print: paged
+  pdf_document: default
+  html_notebook: default
+---
+```
+
 ## WHAT ABOUT THE OTHER LANGUAGES?
@@
 # A RESEARCHER'S WORKFLOW
+
+Now that we have thoroughly explored the what of R Markdown, let us
+now delve into _why_ this tool is useful for researchers.
 

### 2019-01-07 paper.Rmd
tweak whitespace presentation

@@
 
-We will focus on R code for now, and come back to the other languages later. R
-Markdown has a handful of noteworthy differences from general Markdown. First,
+R Markdown has a handful of noteworthy differences from general Markdown. First,
 every R Markdown document begins with a YAML header:
@@
 document. The full documentation on what options are available is [here](todo),
-but the primary elements are the document's Title and Output type. The output type
-has many options, and can actually support multiple options at the same time.
+but the primary elements are the document's Title and Output type. The output
+type has many options, and can actually support multiple options at the same
+time.
 
@@
 
-Code chunks are much like inline code, but they stand alone and are more useful for
-lengthier code blocks that do complex logic, setup, value preparation, plotting, or
-other output generation. It is worth noting that code chunks are [very flexible](todo)
-in their options and defaults, and the number of options can sometimes be overwhelming.
+Code chunks are much like inline code, but they stand alone and are more useful
+for lengthier code blocks that do complex logic, setup, value preparation,
+plotting, or other output generation. It is worth noting that code chunks are
+[very flexible](todo) in their options and defaults, and the number of options
+can sometimes be overwhelming.
 
 Embedded images and other resources are included into R Markdown much like they
-would be in an HTML page. The syntax looks like `![Text alias](path/to/resource)`.
+would be in an HTML page. The syntax looks like `![Text
+alias](path/to/resource)`.
 
 Once the document is written, it is passed through the `knitr` package to "knit"
-the R Markdown document into its final output(s). This creates a new
-/ fresh R context to render the document and thereby helps ensure
-reproducibility.
+the R Markdown document into its final output(s). This creates a new / fresh R
+context to render the document and thereby helps ensure reproducibility.
 
@@
 
-The most noteworthy difference about R Notebooks versus plain R Markdown is that
-they are somewhat _stateful_. This means that they cache and display output as
-it is generated, making it easier to interactively and iteratively build a
+There can be [many reasons](todo) that a user might prefer R Notebooks, but the
+most noteworthy difference between R Notebooks and plain R Markdown is that R
+Notebooks are somewhat _stateful_. This means that they cache and display output
+as it is generated, making it easier to interactively and iteratively build a
 narrative without having to re-render the document each time you have made a
@@
 As a result, R Notebooks can be thought about as good for interactive
-exploration or discussion among developers, but moving to an R Markdown format
-is generally preferred once the report has stabilized or is being shared with an
-end user.
+exploration or discussion among developers, but moving to a standard R Markdown
+format is generally preferred once the report has stabilized or is being shared
+with an end user.
 
@@
 
-## WHAT ABOUT THE OTHER LANGUAGES?
+## PARAMETERIZED R MARKDOWN
 
-## PARAMETERIZED R MARKDOWN
+A little known feature of R Markdown is the ability to add _parameters_.
+Parameterized R Markdown is defined by adding parameters to the YAML header,
+which will then be utilized within the rendering of the document.
 

### 2019-01-07 ex/ex-notebook.Rmd
add example notebook

---
title: "R Notebook"
output:
  html_document:
    df_print: paged
  pdf_document: default
  html_notebook: default
---

This is an [R Markdown](http://rmarkdown.rstudio.com) Notebook. When you execute code within the notebook, the results appear beneath the code.

Try executing this chunk by clicking the *Run* button within the chunk or by placing your cursor inside it and pressing *Cmd+Shift+Enter*.

```{r}
plot(cars)
```

Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Cmd+Option+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Cmd+Shift+K* to preview the HTML file).

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.

### 2019-01-07 paper.Rmd
add parameterized r markdown section

@@
 A little known feature of R Markdown is the ability to add _parameters_.
-Parameterized R Markdown is defined by adding parameters to the YAML header,
-which will then be utilized within the rendering of the document.
+Parameterized R Markdown is defined by the addition of parameters to the YAML
+header, which will then be utilized within the rendering of the document.
+
+For instance, a YAML header like the following might be used to define the
+report dates:
+
+```yaml
+---
+title: "R Markdown - Parameterized"
+output: html_document
+params:
+  start_date: !r Sys.Date() - 10
+  end_date: !r Sys.Date()
+---
+```
+
+This allows these values to be accessible in code using `params$start_date` and
+`params$end_date`. It essentially makes the R Markdown document a _parameterized
+executable_ or function, which can then be called using `rmarkdown::render()`.
+For example, the following executes this R Markdown report with non-default
+parameter selections:
+
+```r
+rmarkdown::render(
+  "./ex/ex-parameterized.Rmd", 
+  params = list(
+    start_date = Sys.Date()-5, 
+    end_date = Sys.Date()
+  )
+)
+```
+
+There is [much more](todo) available to explore in Parameterized R Markdown,
+like an easy-to-use [user interface](todo) for selecting values. You can also
+host Parameterized R Markdown in a publishing platform like RStudio Connect that
+exposes a user interface to end users of a report.
 

### 2019-01-07 paper.Rmd
start section on a researcher's workflow

@@
 
+## A COMMON PROBLEM
+
+Let us imagine a researcher who has done a fair amount of data analysis,
+exploration, and deduction to come to a set of results, and that programming
+was involved in this researcher's endeavor. When the researcher has finished,
+she will likely take the results and use them to build a paper, report, presentation,
+or other output which she will use to share the output with colleagues or 
+other researchers.
+
+This process is tedious and error prone: 
+ - moving the results to another medium
+ - copying values, data, plots, images, etc.
+ - injecting values into prose
+
+Moreover, let us imagine that someone wants to reproduce her work, or that
+another set of surveys, experiments, or clinical trials are added to her
+dataset. Not only must she re-execute her code, she must also duplicate all of
+the transmission of information into her manual "publishing" process. If any
+data or information is accidentally "left in place" from her first pass, her
+results will be un-reproducible and incorrect.
+
+## AN ELEGANT SOLUTION
+
+Let us now imagine the same researcher begins her programming process by writing
+her prose and code in a single R Markdown document. In-line values in the
+document, plots, images, and other output is all generated dynamically based
+only on the input data and runtime ecosystem (packages, etc.).
+
+As a result, when another researcher wants to reproduce her results or when she
+wants to re-execute the paper on new data, the workload becomes "Click Knit."
+Further, our researcher can trust that the report accurately reflects the input
+data, since no human "copy/paste" or other editorial process has taken place.
+
+## WORKFLOW FOR DATA DRIVEN RESEARCH
+
 # OTHER TOOLS

### 2019-01-07 paper.Rmd
make progress on workflow section

@@
 This process is tedious and error prone: 
- - moving the results to another medium
+ - moving the results to another medium (from script output to a PowerPoint)
  - copying values, data, plots, images, etc.
- - injecting values into prose
+ - injecting values into prose (i.e. There are 45 rows in this dataset)
 
@@
 
-# OTHER TOOLS
+This problem and solution can be reduced to a paradigm of _best practices_ in
+which reproducible, data-driven research is greatly simplified by R Markdown.
+There are many [much more thorough](todo) articles on the topic, but we will
+hit the highlights.
+
+- Define input datasets, resources, or other dependencies
+- Manage R packages in a [reproducible fashion](todo)
+- Write narrative and code together into an R Markdown document
+- Ensure that the narrative is programmatically dependent on the data
+- Configure settings so that the R Markdown document produces final output
+
+## AN EXAMPLE
 
@@
 - RStudio documentation, etc
+
+# OTHER TOOLS
+
+## INCLUDING PYTHON
+
+## CUSTOMIZED EMAILS
 

### 2019-01-07 ex/ex-research.Rmd
add silly example research paper

---
title: "Iris Research"
author: "Cole Arendt"
date: "1/6/2019"
output: pdf_document
---

```{r setup, include=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = FALSE)

input_data <- readRDS("iris_altered.rds")

library(dplyr)
library(tidyr)
library(ggplot2)
```

## Iris Data

This dataset is quite an interesting one. It has `r ncol(input_data)` columns
and `r nrow(input_data)` rows. We can look at a plot of the data to get a better
understanding of it:

```{r plot}
ggplot(input_data, aes(Sepal.Length, Sepal.Width)) +
  geom_point() +
  facet_grid(~Species)
```

Further, we can dig into the various groups of the data to see more detail:

```{r explore_no_output, include=FALSE}
mean_detail <- input_data %>%
  group_by(Species) %>%
  summarize_all(mean) %>%
  rename_at(vars(-Species), function(x){paste0("mean_",x)})
median_detail <- input_data %>%
  group_by(Species) %>%
  summarize_all(median) %>%
  rename_at(vars(-Species), function(x){paste0("median_",x)})

all_detail <- mean_detail %>%
  left_join(median_detail, by = c("Species"))

all_detail_tall <- all_detail %>%
  gather(key = "metric", value = "value", -Species) %>%
  spread(key = "Species", value = "value")
```

```{r explore_output}
knitr::kable(all_detail_tall)
```

### 2019-01-07 paper.Rmd
add git reference
add some links add author and date

--- 2019-01-07 paper.Rmd
@@
 title: "R Markdown for Reproducible Research"
+author: "Cole Arendt"
+date: "`r Sys.Date()`"
 output: pdf_document
@@
 
-There can be [many reasons](todo) that a user might prefer R Notebooks, but the
-most noteworthy difference between R Notebooks and plain R Markdown is that R
-Notebooks are somewhat _stateful_. This means that they cache and display output
-as it is generated, making it easier to interactively and iteratively build a
-narrative without having to re-render the document each time you have made a
-change.
+There can be [many
+reasons](https://rviews.rstudio.com/2017/03/15/why-i-love-r-notebooks/) that a
+user might prefer R Notebooks, but the most noteworthy difference between R
+Notebooks and plain R Markdown is that R Notebooks are somewhat _stateful_. This
+means that they cache and display output as it is generated, making it easier to
+interactively and iteratively build a narrative without having to re-render the
+document each time you have made a change.
 
@@
 
-There is [much more](todo) available to explore in Parameterized R Markdown,
-like an easy-to-use [user interface](todo) for selecting values. You can also
-host Parameterized R Markdown in a publishing platform like RStudio Connect that
-exposes a user interface to end users of a report.
+There is [much
+more](https://bookdown.org/yihui/rmarkdown/parameterized-reports.html) available
+to explore in Parameterized R Markdown, like an easy-to-use [user
+interface](https://bookdown.org/yihui/rmarkdown/params-knit.html#the-interactive-user-interface)
+for selecting values. You can also host Parameterized R Markdown in a publishing
+platform like RStudio Connect that [exposes a user interface to end
+users](https://docs.rstudio.com/connect/user/param-rmarkdown.html#publish-param-rmarkdown)
+of a report.
 
@@
 This process is tedious and error prone: 
- - moving the results to another medium
+ - moving the results to another medium (e.g. from script output to a PowerPoint)
  - copying values, data, plots, images, etc.
- - injecting values into prose
+ - injecting values into prose (e.g. There are 45 rows in this dataset)
 
@@
 data or information is accidentally "left in place" from her first pass, her
-results will be un-reproducible and incorrect.
+published results will be un-reproducible and incorrect.
 
@@
 
-# OTHER TOOLS
+This problem and solution can be reduced to a paradigm of _best practices_ in
+which reproducible, data-driven research is greatly simplified by R Markdown.
+There are many [much more thorough](https://rpubs.com/marschmi/105639) articles
+on the topic, but we will focus on the highlights.
+
+- Define input datasets, resources, or other dependencies
+- Manage R packages in a [reproducible
+fashion](https://rviews.rstudio.com/2018/01/18/package-management-for-reproducible-r-code/)
+- Write narrative and code together into an R Markdown document
+- Ensure that the narrative is programmatically dependent on the data
+- Configure settings so that the R Markdown document produces final output
+
+## AN EXAMPLE
 
@@
 
+# OTHER TOOLS
+
+## INCLUDING PYTHON
+
+## CUSTOMIZED EMAILS
+
+## VERSION CONTROL AND GIT
+
 # CONCLUSION
@@
 - http://shop.oreilly.com/product/0636920051435.do
-- https://rviews.rstudio.com/2017/03/15/why-i-love-r-notebooks/
 - https://rviews.rstudio.com/2018/11/01/r-markdown-a-better-approach/
 - https://rviews.rstudio.com/2017/01/25/r-markdown-for-the-enterprise/
-- https://rviews.rstudio.com/2018/01/18/package-management-for-reproducible-r-code/
 - https://en.wikipedia.org/wiki/Markdown

### 2019-01-25 ex/ex-research.Rmd
change dataset back to the basic iris data
parameterize the research
add closing blurb

@@
 output: pdf_document
+params:
+  input_data: "iris.rds"
 ---
@@
 
-input_data <- readRDS("iris_altered.rds")
+input_data <- readRDS(params$input_data)
 
@@
 ```{r explore_output}
+by_med_sepal_length <- all_detail %>% arrange(desc(median_Sepal.Length)) %>% pull(Species)
+
 knitr::kable(all_detail_tall)
 ```
+
+Did you notice that `r by_med_sepal_length %>% .[[1]] %>% as.character()` has
+the highest median Sepal.Length? And `r by_med_sepal_length %>%
+.[[length(by_med_sepal_length)]] %>% as.character()` has the lowest.

### 2019-01-25 paper.Rmd
remove symbol
add clarification about needing R installed

@@
 If Markdown is a lightweight markup language and text formatting syntax, R
-Markdown must be Markdown + R code, right? Yes, with a catch. R Markdown
+Markdown must be Markdown plus R code, right? Yes, with a catch. R Markdown
 actually supports _many_ programming languages and output types. It got its
@@
 users](https://docs.rstudio.com/connect/user/param-rmarkdown.html#publish-param-rmarkdown)
-of a report.
+of a report without needing them to install R.
 

### 2019-01-25 paper.Rmd
add example and python blurb

@@
 
+Let us imagine that we are beginning a project where we explore the classic
+[`iris` dataset](todo). We might approach the problem by first specifying the
+`iris` dataset as our only input. To ensure auditability and reproducibility, I
+might save and [open source the dataset](todo) as well as specify it as a parameter
+for my research in the YAML header.
+
+```
+---
+title: "Iris Research"
+author: "Cole Arendt"
+date: "1/6/2019"
+output: pdf_document
+params:
+  input_data: "iris.rds"
+---
+```
+
+Now, I need R packages to make my work more efficient. I have chosen `dplyr`,
+`tidyr`, and `ggplot2`. Though I reference these packages in my code, I should
+also keep track of which versions of the package I am using. For this task, I
+might use `packrat::.snapshotImpl(".", snapshot.sources = FALSE)`.
+
+Finally, we come to my actual research. As I explore this dataset, I must ensure
+that my results are programmatic. This means plots, values, and prose should
+be dependent only on my input dataset.
+
+When complete, I will be able to generate my research paper with one line of
+code (or one click):
+
+```r
+rmarkdown::render("./ex/ex-research.Rmd")
+```
+
+If I received updated, corrected, or otherwise different data (i.e. "iris_altered.rds"),
+changing my single input parameter and re-rendering the report will reliably rebuild my
+analysis off of different data. Further, any other researcher can take my document and R
+packages to reproduce or validate my results. Feel free to give it a shot yourself!
+
 # TIPS AND TRICKS
 
-- Code chunk options
-- Knit options
-- URLs and internal links
-- Table of Contents
-- Table Output
-- RStudio documentation, etc
+## CODE CHUNK OPTIONS
+
+Every code chunk has _many_ options. These are specified
+
+## KNIT OPTIONS
+
+## URLS AND INTERNAL LINKS
+
+## TABLE OF CONTENTS
+
+## TABLE OUTPUT
+
+## RSTUDIO DOCUMENTATION
 
@@
 ## INCLUDING PYTHON
+
+Python is a general purpose programming language that is increasingly
+common in data science. It has great utility in many regards and many
+useful modules (akin to packages) that can be useful to a researcher.
+
+R Markdown, despite its name, is actually remarkably language agnostic. You
+can include code chunks in other languages (like Python, bash, Stan, and SQL) in
+your R Markdown documents.
+
+Python, in particular, has recently improved support in R Markdown and the RStudio IDE.
+Specifically, the [`reticulate`](todo) package makes it possible to specify the python
+version and conda / virtual environment, as well as persist objects across code chunks. 
+There is also native support for including python plots and other output in your
+rendered R Markdown document. There is much written already about this work flow. If
+interested, the following articles are helpful:
+
+- todo
+- todo
 

### 2019-01-25 paper.Rmd
add notes on code chunks, rstudio documentation, customized emails and git

@@
 
-Every code chunk has _many_ options. These are specified
+Every code chunk has options that control its behavior. These are specified in
+the header above the code chunk:
+
+````
+```{r my_chunk_name, my_option="one", my_other_option="two"}
+r_code <- do_stuff(to_inputs)
+```
+````
+
+While there are [_many_ such options](todo), some of the most ubiquitous are:
+
+- `echo` - whether to "echo" the executed code in the output document
+- `include` - whether to include output of the chunk in the output document
+- `message` - whether to include messages from the chunk in the output document
+
+Further, it is advisable to remember that the defaults can be set with code like:
+```r
+knitr::opts_chunk$set(echo = FALSE)
+```
 
@@
 ## RSTUDIO DOCUMENTATION
+
+RStudio has adopted R Markdown as the standard for some of its most important
+communication. If you are interested in examples, we
+
+- host our [company blog](todo) and [community blog](todo) using [blogdown](todo)
+- write (most) [product documentation](https://docs.rstudio.com/) using [bookdown](todo)
+- build [many](todo) [popular](todo) [websites](todo) using [pkgdown](todo)
+- write presentations and papers like this one using R Markdown
+
+We like to think that the sky is the limit when it comes to the reproducible
+results that can be achieved using R Markdown, and are always excited to see
+the very cool things that the open source community builds using this tool.
 
@@
 
+While emails may not be very relevant to the _researcher's_ task, per se, they
+are a common means of communication for many supportive functions. R Markdown
+reports have what is called [_output metadata_](todo) that can be consumed by
+external services. This, combined with packages like [`blastula`](todo) that
+enable customized email bodies, can facilitate programmatically generating
+reports, documents, or presentations that are programmatically delivered in a
+customizable and personalizable fashion to the consumers of such content.
+
+RStudio Connect is an example of a platform designed to leverage this tooling
+for this end, and there are [several examples]() to explore if interested.
+
 ## VERSION CONTROL AND GIT
+
+Being able to reproduce a single version of one's research is fantastic... but
+what if you could reproduce _every_ version of one's research? That is the sort
+of magic that "Version Control" provides.
+
+`git` is the most popular version control tool, followed by `subversion`. These
+tools are oriented around software engineering and tracking / "diffing" changes
+in source documents (mostly code) over time.
+
+Since our research in this workflow is entirely driven by data along the highway
+of code, it makes sense to spend some time getting used to these tools. If we
+make version control a part of our process, we get reproducible research up to a
+moment in time and easy auditability of code differences since R Markdown
+consists of plaintext documents that are easily compared over time.
+
+There are several helpful tutorials and books on the topic, principally [Happy
+Git with R](todo), and even a built-in `git` pane within the RStudio IDE to simplify
+some operations (todo).
+
+![The RStudio IDE `git` pane with new commits and a modified paper](./img/git-ide.png)
 

### 2019-01-25 paper.Rmd
add (too poetic?) conclusion paragraph

@@
 
+R Markdown is a remarkably powerful tool in the researcher's tool belt. Given a
+community of powerful minds contributing to its future, it will only improve in
+its ability to reliably reproduce researcher's results with minimal effort. I
+imagine a world where reporting standards (like the PhUSE paper formatting) are
+defined in an R Markdown template that any researcher or contributor can easily
+use or extend as needed. It is the power of an open source community building
+quality and reproducible tools that make such a future possible.
+
 ### Resources

### 2019-01-25 paper.Rmd
add urls and table of contents

@@
 
-````
-```{r my_chunk_name, my_option="one", my_other_option="two"}
-r_code <- do_stuff(to_inputs)
+```r
+{r my_chunk_name, my_option="one", my_other_option="two"}
 ```
-````
 
@@
 
-## URLS AND INTERNAL LINKS
+"Knitting" is the process of taking an R Markdown document and generating its
+final output. As a result, options passed to the knit process are "global" (i.e. 
+affect the entire document). Most knit options are defined in the YAML header, and
+common options include:
+
+- templates, formatting, or other output styles
+- parameters
+- output document type
+-
 
@@
 
-## TABLE OUTPUT
+If your document is very long, a table of contents is very helpful. Unfortunately,
+a table of contents is not included in the PhUSE style guide, so we excluded it
+from this paper. However, you can easily include a table of contents by adding
+a section like the following to your YAML header.
+
+```yaml
+output:
+  html_document:
+    toc: true
+```
+
+```yaml
+output:
+  pdf_document:
+    toc: true
+```
+
+## URLS AND INTERNAL LINKS
+
+Often within a research article, report, presentation, or other work, we are standing
+on the shoulders of giants. As a result, it is very important that we appropriately reference
+their work. External URLs are standard markdown format `[link text](link url)`, and these can
+be used to reference internal headers as well. However, for pdf and html document compatibility,
+the following is recommended for internal links:
+
+- Label a header with something like `# MY HEADER {#the-header}`
+- Reference the header using `\@ref(the-header)`
 

### 2019-01-25 paper.Rmd
misc cleanup and addition of some todo items

@@
 almost exclusively in browsers, Markdown can be relevant for the web, print, and
-even computer terminals.
+even computer terminals (todo image).
 
@@
 
-When creating a new document in the RStudio IDE, you might notice an option
+When creating a new document in the RStudio IDE (todo link to figure), you might notice an option
 to create a "R Notebook" or "R Markdown Notebook." This is offered in addition
@@
 - output document type
--
 

### 2019-01-25 paper.Rmd
add another todo
fix small style issues

@@
 
-The researcher's task is to uncover new things; challenge conventional wisdom;
-explore uncharted territory; and confirm or dismiss bias with facts. In order to
+The researcher's task is to uncover new things, challenge conventional wisdom,
+explore uncharted territory, and confirm or dismiss bias with facts. In order to
 accomplish such feats, it is important that a researcher's results be
@@
 for industries like the Pharmaceutical Industry, reproducibility can be a mandate and 
-legal requirement for valid results. As a result, tools like R Markdown that help ensure
-reproducible results are valuable tools in a researcher's toolkit.
+legal requirement for valid results. As a result, software like R Markdown that helps ensure
+reproducible results is a valuable tool in a researcher's toolkit.
 
@@
 
+(todo reference figures, reference bibliography items)
+
 ## RSTUDIO DOCUMENTATION

### 2019-01-25 paper.Rmd
add stylistic fixes

@@
 actually supports _many_ programming languages and output types. It got its
-start with R, and R is the primary language engine, but R is an interface
+start with R, and R is the primary language engine. However, R is an interface
 language. This means that R has been (from its outset) commonly used to "wrap"
@@
 
-It is worth noting that this output can be **problematic for reproducibility**,
-though, since you may have unknowingly persisted an output (state) that is no
-longer relevant or reproducible.
+It is worth noting that this output can be **problematic for reproducibility.**
+Because output is cached, there is _state_ introduced that might unknowingly
+persist when it is no longer relevant or reproducible.
 
@@
 This process is tedious and error prone:
+
  - moving the results to another medium (e.g. from script output to a PowerPoint)
@@
 which reproducible, data-driven research is greatly simplified by R Markdown.
-There are many [much more thorough](https://rpubs.com/marschmi/105639) articles
+There are certainly [more thorough](https://rpubs.com/marschmi/105639) articles
 on the topic, but we will focus on the highlights.
@@
 results that can be achieved using R Markdown, and are always excited to see
-the very cool things that the open source community builds using this tool.
+the impressive works that the open source community builds using this tool.
 
@@
 make version control a part of our process, we get reproducible research up to a
-moment in time and easy auditability of code differences since R Markdown
-consists of plaintext documents that are easily compared over time.
+moment in time. Further, since R Markdown consists of plaintext documents, we
+acquire easy auditability of code that can be compared over time.
 

### 2019-01-25 paper.Rmd
fix figure reference

@@
 date: "`r Sys.Date()`"
-output: pdf_document
+output: bookdown::pdf_document2
 ---
@@
 
-When creating a new document in the RStudio IDE (todo link to figure), you might notice an option
+When creating a new document in the RStudio IDE (see Figure \@ref(file-dropdown)), you might notice an option
 to create a "R Notebook" or "R Markdown Notebook." This is offered in addition
@@
 ![RStudio IDE New File Dropdown showing R Markdown and R Notebook
-options](./img/new-file-dropdown.png)
+options](./img/new-file-dropdown.png){#file-dropdown}
 

### 2019-01-25 paper.Rmd
fix many links remove all todo items

@@
 almost exclusively in browsers, Markdown can be relevant for the web, print, and
-even computer terminals (todo image).
+even computer terminals (see Figure \@ref(terminal-markdown)).
+
+![Markdown with syntax highlighting in a computer terminal](./img/terminal-markdown.png){#terminal-markdown}
 
@@
 The YAML header uses YAML syntax to declare all sorts of information about the
-document. The full documentation on what options are available is [here](todo),
-but the primary elements are the document's Title and Output type. The output
-type has many options, and can actually support multiple options at the same
-time.
+document. The full documentation on what options are available is
+[here](https://www.rstudio.com/wp-content/uploads/2015/03/rmarkdown-reference.pdf)
+or [here](https://bookdown.org/yihui/rmarkdown/), but the primary elements are
+the document's Title and Output type. The output type has many options, and can
+actually support multiple options at the same time.
 
@@
 plotting, or other output generation. It is worth noting that code chunks are
-[very flexible](todo) in their options and defaults, and the number of options
-can sometimes be overwhelming.
+[very
+flexible](https://www.rstudio.com/wp-content/uploads/2015/03/rmarkdown-reference.pdf)
+in their options and defaults, and the number of options can sometimes be
+overwhelming.
 
@@
 Let us imagine that we are beginning a project where we explore the classic
-[`iris` dataset](todo). We might approach the problem by first specifying the
-`iris` dataset as our only input. To ensure auditability and reproducibility, I
-might save and [open source the dataset](todo) as well as specify it as a parameter
-for my research in the YAML header.
+[`iris`
+dataset](https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/iris.html).
+We might approach the problem by first specifying the `iris` dataset as our only
+input. To ensure auditability and reproducibility, I might save and [open source
+the dataset](http://opendatahandbook.org/guide/en/why-open-data/) as well as
+specify it as a parameter for my research in the YAML header.
 
@@
 
-While there are [_many_ such options](todo), some of the most ubiquitous are:
+While there are [_many_ such
+options](https://www.rstudio.com/wp-content/uploads/2015/03/rmarkdown-reference.pdf),
+some of the most ubiquitous are:
 
@@
 
-(todo reference figures, reference bibliography items)
+Referencing images can be done in a similar fashion. It unfortunately requires
+using a different output format for PDFs (`bookdown::pdf_document2`), but
+can be done using:
+
+- Label the image with `![Image alt text](./image/path){#the-label}`
+- Reference the image using `Figure \@ref(the-label)`
+- Images [built in a code
+chunk](https://bookdown.org/yihui/bookdown/figures.html) can also be referenced
+with `\@ref(fig:code-chunk-name)`
 
@@
 
-- host our [company blog](todo) and [community blog](todo) using [blogdown](todo)
-- write (most) [product documentation](https://docs.rstudio.com/) using [bookdown](todo)
-- build [many](todo) [popular](todo) [websites](todo) using [pkgdown](todo)
+- host our [company blog](https://blog.rstudio.com/) and [community
+blog](https://rviews.rstudio.com/) using
+[blogdown](https://bookdown.org/yihui/blogdown/)
+- write (most) [product documentation](https://docs.rstudio.com/) using
+[bookdown](https://bookdown.org/yihui/bookdown/)
+- build [many](https://ggplot2.tidyverse.org/)
+[popular](https://db.rstudio.com/) [websites](http://shiny.rstudio.com/) using
+[pkgdown](https://pkgdown.r-lib.org/)
 - write presentations and papers like this one using R Markdown
@@
 
-Python, in particular, has recently improved support in R Markdown and the RStudio IDE.
-Specifically, the [`reticulate`](todo) package makes it possible to specify the python
-version and conda / virtual environment, as well as persist objects across code chunks. 
+Python, in particular, has recently improved support in R Markdown and the
+RStudio IDE. Specifically, the
+[`reticulate`](https://rstudio.github.io/reticulate/) package makes it possible
+to specify the python version and conda / virtual environment, as well as
+persist objects across code chunks.
+
 There is also native support for including python plots and other output in your
-rendered R Markdown document. There is much written already about this work flow. If
-interested, the following articles are helpful:
+rendered R Markdown document. There is much written already about this work
+flow. If interested, the following articles are helpful:
 
-- todo
-- todo
+- [RStudio IDE
+features](https://blog.rstudio.com/2018/10/09/rstudio-1-2-preview-reticulated-python/)
+- [Calling Python from
+R](https://rstudio.github.io/reticulate/articles/calling_python.html)
+- [An example Shiny
+application](https://rviews.rstudio.com/2018/04/17/reticulated-shiny/)
 
@@
 are a common means of communication for many supportive functions. R Markdown
-reports have what is called [_output metadata_](todo) that can be consumed by
-external services. This, combined with packages like [`blastula`](todo) that
-enable customized email bodies, can facilitate programmatically generating
-reports, documents, or presentations that are programmatically delivered in a
-customizable and personalizable fashion to the consumers of such content.
+reports have what is called [_output
+metadata_](https://docs.rstudio.com/connect/1.7.0/user/r-markdown.html#r-markdown-output-metadata)
+that can be consumed by external services. This, combined with packages like
+[`blastula`](https://github.com/rich-iannone/blastula) that enable customized
+email bodies, can facilitate programmatically generating reports, documents, or
+presentations that are programmatically delivered in a customizable and
+personalizable fashion to the consumers of such content.
 
 RStudio Connect is an example of a platform designed to leverage this tooling
-for this end, and there are [several examples]() to explore if interested.
+for this end, and there are [several
+examples](https://blog.rstudio.com/2018/07/26/rstudio-connect-1-6-6-custom-emails/)
+to explore if interested.
 
@@
 There are several helpful tutorials and books on the topic, principally [Happy
-Git with R](todo), and even a built-in `git` pane within the RStudio IDE to simplify
-some operations (todo).
+Git with R](https://happygitwithr.com/), and even a built-in `git` pane within the RStudio IDE to simplify
+some operations (see Figure \@ref(git-ide)).
 
-![The RStudio IDE `git` pane with new commits and a modified paper](./img/git-ide.png)
+![The RStudio IDE `git` pane with new commits and a modified paper](./img/git-ide.png){#git-ide}
 
@@
 
-- https://rpubs.com/marschmi/105639
 - http://ropensci.github.io/reproducibility-guide/sections/tools/
@@
 - https://www.oreilly.com/library/view/the-art-of/9781593273842/ch15.html
-
-### TODO
-- Link for YAML syntax
-- Link for YAML header docs
-- Resource on R as an interface language
-- link to figures / images
-- start managing references better...

### 2019-01-25 paper.Rmd
formalize the resources section

@@
 
-The story of R Markdown begins with Markdown, a markup language and formatting
+The story of R Markdown begins with
+[Markdown](https://en.wikipedia.org/wiki/Markdown), a markup language and
+formatting
 syntax designed to be:
@@
 
-### Resources
+# RESOURCES
+
+The source and examples from this paper are available on GitHub:
+
+https://github.com/colearendt/rmarkdown-for-reproducible-research
+
+There are also many other interesting resources on this topic
+which we did not have the opportunity to include:
 
@@
 - https://rviews.rstudio.com/2017/01/25/r-markdown-for-the-enterprise/
-- https://en.wikipedia.org/wiki/Markdown
-- https://www.oreilly.com/library/view/the-art-of/9781593273842/ch15.html
